---
title: "Taxonomic analysis"
author: "Yanxian Li<br><small>Department of Paraclinical Sciences<br>Faculty of Veterinary Medicine<br>Norwegian University of Life Sciences</small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    toc_float: 
      collapsed: true
    code_folding: show
    theme: cerulean
    self_contained: true
  pdf_document: 
    latex_engine: xelatex
  word_document: default  
---

```{r style, echo = FALSE, message = FALSE, warning = FALSE}
require(knitr)
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE, cache = FALSE)
```

`r Hmisc::hidingTOC(buttonLabel = "Outline (hide)", tocSide = "right",  buttonSide = "right")`

# Getting ready

Load packages.

```{r load-package}
library(here) # A Simpler Way to Find Your Files, CRAN
library(tidyverse) # Easily Install and Load the 'Tidyverse', CRAN
library(MicrobeR) # Handy functions for microbiome analysis in R, [github::jbisanz/MicrobeR] 
library(microbiome) # Microbiome Analytics, Bioconductor
library(speedyseq) # Faster implementations of phyloseq functions, [github::mikemc/speedyseq] 
library(RColorBrewer) # ColorBrewer Palettes, CRAN
library(patchwork) # The Composer of Plots, CRAN
library(flextable) # Functions for Tabular Reporting, CRAN
library(officer) # Manipulation of Microsoft Word and PowerPoint Documents, CRAN
```

Load data.

```{r}
# load phyloseq object
ps <- readRDS(here("data/intermediate/qiime2R/phyloseq.rds"))

# filter samples
ps <- subset_samples(ps, Population == "Wild") 

# change feature names for easy display
indx <- formatC(1:ntaxa(ps), width = nchar(ntaxa(ps)), format = "d", flag = "0")
taxa_names(ps) <- paste0("ASV", indx)
```

Extract metadata, feature table and taxonomy 

```{r}
# metadata
mtd <- data.frame(sample_data(ps), check.names = FALSE) 

# feature table
asv_tbl <- as.data.frame(otu_table(ps)) 

# taxonomy
txnm_tbl <- tax_table(ps) %>% as("matrix") %>% as.data.frame()
```

# Taxa barplot

## Summarize data

Get top 10 most abundant taxa at genus level.

```{r}
taxa_tab <- Summarize.Taxa(asv_tbl, txnm_tbl)$Genus %>% Make.Percent() 
taxa_tab <- taxa_tab[order(rowMeans(taxa_tab), decreasing = T), ]
Others <- colSums(taxa_tab[11:nrow(taxa_tab), ])
taxa_tab <- rbind(taxa_tab[1:10, ], Others)
```

Tidy taxa table for making stacked bar plots.

```{r}
taxa_tab <- as.data.frame(taxa_tab) %>%
  rownames_to_column("Taxa") %>%
  separate(
    Taxa, 
    sep = ";", 
    c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")) %>% 
  mutate(
    Phylum = ifelse(
      is.na(Phylum)|Phylum == "NA"|grepl("uncultured|Ambiguous|metagenome", Phylum), 
      Kingdom, 
      Phylum),
    Class = ifelse(
      is.na(Class)|Class == "NA"|grepl("uncultured|Ambiguous|metagenome", Class), 
      Phylum, 
      Class),
    Order = ifelse(
      is.na(Order)|Order == "NA"|grepl("uncultured|Ambiguous|metagenome", Order), 
      Class, 
      Order),
    Family = ifelse(
      is.na(Family)|Family == "NA"|grepl("uncultured|Ambiguous|metagenome", Family), 
      Order, 
      Family),
    Genus = ifelse(
      is.na(Genus)|Genus == "NA"|grepl("uncultured|Ambiguous|metagenome", Genus), 
      Family, 
      Genus)) %>%
  select(-Kingdom, -(Class:Family)) %>%
  mutate(
    Phylum = gsub("p__", "", Phylum),
    Phylum = factor(Phylum, levels = rev(unique(Phylum))),
    Genus = gsub("g__", "", Genus),
    Genus = factor(Genus, levels = rev(unique(Genus)))
    ) %>%
  arrange(Phylum, Genus) %>%
  mutate(Genus = factor(Genus, levels = unique(Genus))) %>%
  pivot_longer(
    cols = -c(Phylum, Genus), 
    names_to = "SampleID", 
    values_to = "Abundance"
    ) %>%
  inner_join(rownames_to_column(mtd, "SampleID"), by = "SampleID")
```

## Make stacked barplot

Define color scheme.

```{r}
# taxa names of the most abundant genera 
top_tax <- filter(taxa_tab, Phylum != "Others") %>% distinct(Phylum, Genus)

# make functions to generate color gradients for genera within each phylum 
getCol <- length(unique(top_tax$Phylum)) %>% # number of phyla 
  brewer.pal(name = "Set2") %>% # base colors of color gradients
  rev() %>%
  map(~c(.x, "white")) %>% # start and end color of color gradients
  map(colorRampPalette) # functions to make color gradients

# generate color gradients 
n <- dplyr::count(top_tax, Phylum) %>% pull(n) # number of genera within each phylum
col <- map2(getCol, n, ~rev(.x(.y + 1)[1:.y])) %>% unlist() # generate color gradients & remove white color

# use grey color for the category "Others"
col <- c("grey", col)
```

Make plot.

```{r}
taxa_bar <- taxa_tab %>%
  ggplot(aes(x = SampleID, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample", y = "Relative abundance (%)") +
  scale_y_continuous(breaks = 0:10*10, expand = c(0,0)) + 
  scale_fill_manual(values = col) +
  facet_wrap(~ Species, scales = "free_x") +
  theme_bw() 

taxa_bar
```

## Modify legend

Modify legend using flextable.

```{r}
# legend to flextable
lgnd_ft <- unique(taxa_tab[ , c("Genus", "Phylum")]) %>%
  mutate(col = 11, .before = Genus) %>% # legend color
  flextable() %>% 
  theme_box() %>%
  highlight(j = "col", color = col, part = "body") %>% # use legend color to highlight text
  color(j = "col", color = col, part = "body") %>% # use legend color as font color to hide the text
  color(j = "col", color = "white", part = "header") %>% # hide text in the table header
  italic(i = 2:11, italic = TRUE, part = "body") %>% # italicize taxa names
  fontsize(size = 16, part = "all") %>% # font size
  width(j = c("col"), width = 0.2) %>% # change the width of legend column
  height_all(height = 1) %>% # cell height
  align(j = c("Genus", "Phylum"), align = "left", part = "all") %>% # left-align text 
  merge_v(j = "Phylum") %>% # merge cells by phylum
  border_remove() %>% # remove borders
  theme_booktabs() %>% # apply table style
  hline(i = c(1, 2, 3, 7, 10), border = fp_border(width = 1.5), part = "body") %>% # draw horizontal lines  
  hline_top(part = "header", border = fp_border(color = "white")) # remove horizontal line in the table header

# convert flextable to raster
lgnd_rst <- as_raster(lgnd_ft, webshot = "webshot2") 

# convert raster to ggplot
lgnd_gg <- ggplot() + 
  theme_void() + 
  annotation_custom(grid::rasterGrob(lgnd_rst)) # top right: y = unit(0.72, "npc")
```

## Figure 1

```{r, fig.width=8, fig.height=5}
taxa_bar + theme(legend.position = "none") + lgnd_gg +
  plot_layout(widths = c(2, 1))

# export plot
ggsave(here("result/figure/perch/Figure 1.tiff"), width = 8, height = 5,
       units = "in", dpi = 300, compression = "lzw")
```

# Core microbiota

## Data preprocessing

```{r}
# subset phyloseq
ps_ls <- list(
  perch = subset_samples(ps, Species == "Nile perch"),
  tilapia = subset_samples(ps, Species == "Nile tilapia")
  )

# data filtering and transformation
ps_ls <- ps_ls %>%
  # remove features with zero count in all samples
  map(~filter_taxa(.x, function(x) sum(x > 0) > 0, TRUE)) %>%
  # total sum scaling
  map(~transform_sample_counts(.x, function(x){x / sum(x)}))
```

## Core microbiota heatmap

### Base plot

```{r, results='hide'}
# base plot
coreHmp <- map(seq_along(ps_ls), function(x){
  plot_core(ps_ls[[x]], 
    plot.type = "heatmap", 
    colours = rev(brewer.pal(5, "RdBu")),
    prevalences = seq(0.1, 1, 0.1), 
    detections = seq(0.0001, 0.01, length = 10),
    min.prevalence = 0.8) 
  }
)

# names of the list elements
names(coreHmp) <- c(
  "Core ASVs in wild Nile perch", 
  "Core ASVs in wild Nile tilapia"
  )  

# modify aesthetics
coreHmp <- map(seq_along(coreHmp), function(x){
  coreHmp[[x]] + 
  scale_fill_gradientn(
    colours = rev(brewer.pal(5, "RdBu")),
    breaks = seq(0, 1, 0.1), 
    labels = formatC(seq(0, 1, 0.1), format = "f", digits = 1), 
    limits = c(0, 1)) +
  labs(
    title = names(coreHmp)[x], 
    x = "Detection threshold (%)", y = "") +
  theme_bw() +
  theme(
    #plot.title.position = "plot",
    #plot.title = element_text(hjust = 0.05)
  )
  }
)  
```

### Format x-axis text

```{r}
# get plot data
coreHmp_dt <- map(coreHmp, ~.x$data) 

# format x axis labels
xlab <- map(
  coreHmp_dt, 
  ~unique(.x$DetectionThreshold) %>% as.character() %>% as.numeric() # factor to numeric
  ) %>%  
  map(~formatC(.x * 100, format = "f", digits = 2))   

# modify x axis labels
coreHmp <- map2(coreHmp, xlab, ~.x + scale_x_discrete(labels = .y)) 
```

### Format y-axis text

```{r}
# ASVs shared between the core microbiome heatmaps
core_shrd <- Reduce(intersect, lapply(coreHmp_dt, "[[", "Taxa")) %>%
  paste0(collapse = "|")

# the taxonomy of core ASVs
coreHmp_dt <- map(coreHmp_dt, function(x){
  txnm_tbl[levels(x$Taxa), ] %>%
  rownames_to_column("Taxa") %>%
  inner_join(x, by = "Taxa") %>%
  # get the best taxonomic annotations for core taxa
  mutate(
    Class = ifelse(
      is.na(Class)|Class == "NA"|grepl("uncultured|Ambiguous|metagenome", Class), 
      Phylum, 
      Class
      ),
    Order = ifelse(
      is.na(Order)|Order == "NA"|grepl("uncultured|Ambiguous|metagenome", Order), 
      Class, 
      Order
      ),
    Family = ifelse(
      is.na(Family)|Family == "NA"|grepl("uncultured|Ambiguous|metagenome", Family), 
      Order, 
      Family
      ),
    Genus = ifelse(
      is.na(Genus)|Genus == "NA"|grepl("uncultured|Ambiguous|metagenome", Genus), 
      Family, 
      Genus)
    ) %>%
  select(Taxa, Genus, DetectionThreshold, Prevalence) %>%
  # italicize taxa names using R expressions
  mutate(
    Taxa = paste0("italic(", Genus, ")", "(", Taxa, ")"),
    # some taxa names contain space; tilde (~) is recognized as "space" in R expressions
    Taxa = gsub("\\s+", "~", Taxa), 
    Taxa = gsub("g__", "", Taxa),
    Taxa = factor(Taxa, levels = unique(Taxa))
    ) %>%
  select(-Genus)
  }
)

# parse the taxonomy of core ASVs as R expression 
ylab <- map(coreHmp_dt, ~parse(text = levels(.x$Taxa))) 

# modify y axis labels and highlight shared core ASVs in red color
coreHmp <- map2(coreHmp, ylab, function(x, y){
  x + scale_y_discrete(labels = y) + 
    theme(
      axis.text.y = element_text(
      colour = ifelse(grepl(core_shrd, levels(x$data$Taxa)), 'red', 'black')
        )
      )
    }
  ) 
```

## Figure 2

```{r, fig.width=13, fig.height=6}
# assemble plots
wrap_plots(coreHmp, nrow = 1, guides = "collect") +
  plot_annotation(tag_levels = "a")

# export plot
ggsave(here("result/figure/perch/Figure 2.tiff"), width = 13, height = 6,
       units = "in", dpi = 300, compression = "lzw")
```

# Session information

```{r session-info}
sessionInfo()
```
