---
title: "Alpha diversity"
author: "Yanxian Li<br><small>Department of Paraclinical Sciences<br>Faculty of Veterinary Medicine<br>Norwegian University of Life Sciences</small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    toc_float: 
      collapsed: true
    code_folding: show
    theme: cerulean
    self_contained: true
  pdf_document: 
    latex_engine: xelatex
  word_document: default  
---

```{r style, echo = FALSE, message = FALSE, warning = FALSE}
require(knitr)
options(width = 120)
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE, cache = FALSE)
```

`r Hmisc::hidingTOC(buttonLabel = "Outline (hide)", tocSide = "right",  buttonSide = "right")`

# Getting ready

Load packages

```{r}
library(here) # A Simpler Way to Find Your Files, CRAN
library(tidyverse) # Easily Install and Load the 'Tidyverse', CRAN
library(patchwork) # The Composer of Plots, CRAN
library(afex) # Analysis of Factorial Experiments, CRAN
library(ggResidpanel) # Panels and Interactive Versions of Diagnostic Plots, CRAN

# Set seed
set.seed(1910)
```

Load data

```{r}
# metadata
mtd <- read_tsv(here("data/metadata.tsv"), comment = "#q2") %>%
  rename(SampleID = "#SampleID")
  
# alpha diversity
alph <- readRDS(here("data/intermediate/qiime2R/perch/alpha_diversity.rds"))
```

Tidy data

```{r}
alph <- map_dfc(alph, ~.x$data) %>% # merge alpha diversity indices
  rownames_to_column("SampleID") %>%
  inner_join(mtd, by = "SampleID") %>%
  rename(
    "Observed ASVs" = observed_features, 
    "Pielou's evenness" = pielou_evenness, 
    "Shannon's index" = shannon_entropy, 
    "Faith's PD" = faith_pd) %>%
  pivot_longer(
    cols = "Pielou's evenness":"Shannon's index", 
    names_to = "alph_indx", 
    values_to = "value") %>%
  mutate(
    alph_indx = factor(
      alph_indx, 
      levels = c(
        "Observed ASVs", "Pielou's evenness", "Shannon's index", "Faith's PD"
        )
      )
    )
```

# Plot alpha diversity

```{r, fig.width=9, fig.height=4}
p <- alph %>% 
  ggplot(aes(x = Species, y = value)) +
  geom_point(aes(shape = Species), color = "Darkgrey", alpha = 0.8, size = 2) +
  stat_summary(
    aes(color = Species, shape = Species),
    fun.data = "mean_sdl", 
    fun.args = list(mult = 1), 
    geom = "pointrange",
    size = 0.8) +
  facet_wrap(~ alph_indx, scales = "free", nrow = 1) +
  scale_y_continuous(
    limits = c(0, NA), 
    expand = expand_scale(mult = c(0, 0.1))
    ) +  
  theme_bw(base_size = 14) +
  scale_color_brewer(palette = "Set2") +
  theme(legend.position = "none")

p
```

# Compare alpha diversity

## Fit linear mixed effects models

Here we use the `mixed` function from the *afex* package to fit linear mixed effects models. The *p*-value of fixed effects is calculated using the Kenward-Roger approximation, which provides the best control against anti-conservative results.

```{r, message=TRUE}
# split data
alph_spl <- split(alph, alph$alph_indx)

# fit model
lme <- lapply(alph_spl, function(x) {
  mixed(
    log1p(value) ~ Species + (1|Site), 
    data = x, 
    method = "KR"
    )
  }
)
```

Model summary.

```{r}
lapply(lme, anova)
```

## Model diagnostics

Here we use the *ggResidpanel* package to produce a panel of plots for residual diagnostics.

```{r, results='hide'}
lapply(
  seq_along(lme), 
  function(x) 
  {
  # extract title
  main <- grid::textGrob(names(lme)[x])
  
  # make residual diagnostic plots
  resid_panel <- resid_panel(lme[[x]]$full_model, plots = "all", qqbands = TRUE)
  
  # assemble plots
  wrap_elements(main) / resid_panel +
    plot_layout(heights = c(1, 10))
  }
)
```

Everything looks fine. As the fixed effect, `Species`, has only 2 levels, we will not do follow-up post-hoc tests. 

# Figure 3

Gather *p*-values.

```{r}
pval <- lme %>%
  map_dfr(~rownames_to_column(.x$anova_table, "term"), .id = "alph_indx") %>%
  select(-c(`num Df`:F)) %>%
  rename(p = "Pr(>F)") %>% 
  mutate(
    # format p values
    p = ifelse(
      p < 0.001, 
      "< 0.001", 
      formatC(p, format = "f", digits = 3)
      ), 
    p = ifelse(
      grepl("<", p), 
      paste("italic(p)", p), 
      paste("italic(p) ==", "`", p, "`") # encasing p values with `` to preserve trailing zeros
    ),
    # start position of horizontal bars on x-axis 
    bar_start = "Nile perch",
    # end position of horizontal bars on x-axis 
    bar_end = "Nile tilapia",
    # x-axis position for p value annotation 
    xpos = 1.5,
    # y-axis position for p values and horizontal bars 
    ypos = c(
      2280, # Observed ASVs 
      0.76, # Pielou's evenness
      8.4,  # Shannon's index 
      310   # Faith's PD
      ) 
    )
```

Add *p*-values to alpha diversity plots.

```{r, fig.width=9, fig.height=4}
p +
  # add horizontal bars
  geom_segment( 
    data = pval,
    aes(x = bar_start, xend = bar_end, y = ypos, yend = ypos)
    ) +
  # add p values 
  geom_text( 
    data = pval,
    aes(x = xpos, y = ypos, label = p),
    parse = TRUE,
    size = 3.5, 
    vjust = -0.1
    ) 

# export plot
ggsave(here("result/figure/perch/Figure 3.tiff"), width = 9, height = 4,
       units = "in", dpi = 300, compression = "lzw")
```

# Session information

```{r session-info}
sessionInfo()
```
